---
title: "Warehouse location with orce"
format: html
editor: visual
cache: false
freeze: false
---

```{r setup, include = FALSE}
library(spopt)
library(tinytable)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

The `orce()` function solves the warehouse location problem: given demand points and candidate facilities, decide which facilities to open, how to assign demand, and how many workers to place at each facility --- all while minimizing total cost.

This vignette compares three approaches to the same problem:

1.  **`cflp()`** --- Capacitated Facility Location: minimizes transport + fixed costs, treating capacity as a hard constraint
2.  **`orce()`** --- Warehouse Location: minimizes transport + fixed + worker costs, with worker count as a decision variable
3.  **`orce::orce()`** --- The reference implementation from the [orce](https://github.com/orce-ibge/orce) package

## A synthetic problem

We create 200 demand points (collection units) and 30 candidate facilities (agencies). Each demand point has a workload (collection days), and each facility has a fixed opening cost and a maximum number of workers.

```{r problem-setup}
library(sf)

set.seed(2024)

n_demand <- 200
n_fac <- 30

# Demand points: scattered across a 500x500 km area
demand <- st_as_sf(data.frame(
  id = seq_len(n_demand),
  x = runif(n_demand, 0, 500),
  y = runif(n_demand, 0, 500),
  workload = as.integer(rpois(n_demand, 15) + 5)
), coords = c("x", "y"))

# Facilities: 30 candidates at random locations
fac_x <- runif(n_fac, 10, 490)
fac_y <- runif(n_fac, 10, 490)
facilities <- st_as_sf(data.frame(
  id = seq_len(n_fac),
  x = fac_x,
  y = fac_y,
  fixed_cost = as.integer(runif(n_fac, 300, 800)),
  max_workers = as.integer(sample(3:8, n_fac, replace = TRUE))
), coords = c("x", "y"))
```

### Cost parameters

We model transport cost as fuel cost for a round trip:

$$\text{transport\_cost}_{ij} = \text{workload}_i \times \frac{2 \times \text{distance}_{ij}}{\text{kml}} \times \text{fuel\_cost}$$

Each worker can handle up to 60 collection days. Worker salary is 1500 per worker.

```{r cost-params}
kml <- 10              # km per liter
fuel_cost <- 6         # cost per liter
worker_capacity <- 60  # max workload per worker
worker_cost <- 1500    # salary per worker

# Distance matrix (Euclidean, treating coordinates as km)
dist_mat <- distance_matrix(demand, facilities)

# Per-unit transport cost: cost of one round trip to facility j from demand i
# (Used by cflp, which multiplies by weight internally)
unit_cost_mat <- 2 * dist_mat / kml * fuel_cost

# Total transport cost per demand-facility pair (includes workload weighting)
# (Used by orce, which does NOT multiply by weight)
cost_mat <- outer(demand$workload, rep(1, n_fac)) * unit_cost_mat
```

## Approach 1: CFLP

`cflp()` minimizes transport + fixed facility costs. It doesn't know about workers --- capacity is simply `max_workers * worker_capacity`. Because worker costs are not in the objective, it may open facilities that are cheap to open but expensive to staff.

```{r cflp-solve}
# CFLP needs a single capacity column
facilities$capacity <- facilities$max_workers * worker_capacity

# cflp() multiplies weight * cost internally, so we pass the per-unit cost
cflp_result <- cflp(
  demand, facilities,
  n_facilities = 0,
  weight_col = "workload",
  capacity_col = "capacity",
  facility_cost_col = "fixed_cost",
  cost_matrix = unit_cost_mat
)

cflp_meta <- attr(cflp_result, "spopt")
cflp_selected <- which(cflp_result$facilities$.selected)
```

## Approach 2: spopt::orce()

`orce()` adds worker costs to the objective. The solver decides how many workers each open facility needs, trading off transport savings against workforce costs.

```{r orce-solve}
orce_result <- orce(
  demand, facilities,
  weight_col = "workload",
  cost_matrix = cost_mat,
  facility_cost_col = "fixed_cost",
  worker_cost = worker_cost,
  worker_capacity = worker_capacity,
  max_workers_col = "max_workers"
)

orce_meta <- attr(orce_result, "spopt")
orce_selected <- which(orce_result$facilities$.selected)
```

Workers per open facility:

```{r orce-workers}
open <- orce_result$facilities[orce_result$facilities$.selected, ]
data.frame(
  facility = open$id,
  workers = open$.workers,
  assigned = open$.n_assigned,
  utilization = round(open$.utilization, 2)
) |> tt()
```

## Approach 3: orce::orce() (reference)

The original orce package solves the same MIP using R's ROI infrastructure. We set up the identical problem to verify our Rust implementation matches.

```{r orce-pkg, eval = requireNamespace("orce", quietly = TRUE)}
library(orce)

uc_ids <- paste0("uc_", seq_len(n_demand))
ag_ids <- paste0("ag_", seq_len(n_fac))

ucs_df <- data.frame(
  uc = uc_ids,
  agencia_codigo = ag_ids[1],
  dias_coleta = demand$workload,
  viagens = 1L,
  data = "2024-01",
  diaria_valor = 0,
  stringsAsFactors = FALSE
)

agencias_df <- data.frame(
  agencia_codigo = ag_ids,
  n_entrevistadores_agencia_max = facilities$max_workers,
  custo_fixo = facilities$fixed_cost,
  diaria_valor = 0,
  stringsAsFactors = FALSE
)

distancias_ucs_df <- expand.grid(
  uc = uc_ids,
  agencia_codigo = ag_ids,
  stringsAsFactors = FALSE
)
distancias_ucs_df$distancia_km <- as.vector(dist_mat)
distancias_ucs_df$duracao_horas <- 0
distancias_ucs_df$diaria_municipio <- FALSE
distancias_ucs_df$diaria_pernoite <- FALSE

orce_pkg_args <- list(
  ucs = ucs_df,
  agencias = agencias_df,
  distancias_ucs = distancias_ucs_df,
  dias_coleta_entrevistador_max = worker_capacity,
  remuneracao_entrevistador = worker_cost,
  n_entrevistadores_min = 1L,
  custo_litro_combustivel = fuel_cost,
  custo_hora_viagem = 0,
  kml = kml,
  dias_treinamento = 0,
  weight_tsp = 0,
  adicional_troca_jurisdicao = 0,
  rel_tol = 0,
  use_cache = FALSE
)

# Integer workers (default)
orce_pkg_time <- system.time({
  orce_pkg_result <- do.call(orce::orce, orce_pkg_args)
})

# Continuous workers (LP relaxation of w[j])
orce_pkg_cont_time <- system.time({
  orce_pkg_cont_result <- do.call(orce::orce,
    c(orce_pkg_args, n_entrevistadores_tipo = "continuous"))
})

orce_pkg_obj <- attr(orce_pkg_result, "valor")
orce_pkg_cont_obj <- attr(orce_pkg_cont_result, "valor")
```

## Comparison

```{r comparison-table}
# CFLP transport cost = objective - fixed costs
cflp_transport <- cflp_meta$objective - sum(facilities$fixed_cost[cflp_selected])
cflp_fixed <- sum(facilities$fixed_cost[cflp_selected])

# Compute CFLP's worker cost post-hoc (not part of its optimization)
cflp_workers_per_fac <- ceiling(
  tapply(demand$workload[cflp_result$demand$.facility],
         cflp_result$demand$.facility,
         sum) / worker_capacity
)
cflp_worker_total <- sum(cflp_workers_per_fac) * worker_cost

has_orce_pkg <- exists("orce_pkg_result")

comparison <- data.frame(
  Model = c("cflp()", "orce()",
            if (has_orce_pkg) c("orce::orce(integer)", "orce::orce(continuous)")),
  Facilities = c(
    sum(cflp_result$facilities$.selected),
    orce_meta$n_selected,
    if (has_orce_pkg) c(
      nrow(orce_pkg_result$resultado_agencias_otimo),
      nrow(orce_pkg_cont_result$resultado_agencias_otimo)
    )
  ),
  Transport = round(c(
    cflp_transport,
    orce_meta$transport_cost,
    if (has_orce_pkg) c(
      sum(orce_pkg_result$resultado_ucs_otimo$custo_deslocamento),
      sum(orce_pkg_cont_result$resultado_ucs_otimo$custo_deslocamento)
    )
  ), 1),
  Fixed = round(c(
    cflp_fixed,
    orce_meta$facility_cost,
    if (has_orce_pkg) c(
      sum(orce_pkg_result$resultado_agencias_otimo$custo_fixo),
      sum(orce_pkg_cont_result$resultado_agencias_otimo$custo_fixo)
    )
  ), 1),
  Workers = round(c(
    cflp_worker_total,
    orce_meta$worker_cost_total,
    if (has_orce_pkg) c(
      sum(orce_pkg_result$resultado_agencias_otimo$custo_total_entrevistadores),
      sum(orce_pkg_cont_result$resultado_agencias_otimo$custo_total_entrevistadores)
    )
  ), 1),
  `Time (s)` = round(c(
    cflp_meta$solve_time,
    orce_meta$solve_time,
    if (has_orce_pkg) c(
      orce_pkg_time[["elapsed"]],
      orce_pkg_cont_time[["elapsed"]]
    )
  ), 3),
  check.names = FALSE
)
comparison$Total <- comparison$Transport + comparison$Fixed + comparison$Workers
comparison |> tt()
```

The key insight: `cflp()` and `orce()` may select different facilities because `orce()` accounts for workforce costs during optimization, while `cflp()` only considers transport and fixed costs. When worker costs are significant (as they often are), the `orce()` solution is more cost-effective overall.

## Iterative location-routing (`weight_tsp`)

The basic `orce()` minimizes a cost matrix that treats each demand--facility pair independently. This can produce assignments where nearby demand points are split across distant facilities, because the MIP has no visibility into routing.

When `weight_tsp > 0`, `orce()` adds an iterative feedback loop:

1.  Solve the ORCE MIP with the current cost matrix.
2.  For each opened facility, solve a TSP tour over its assigned demands.
3.  For every (demand, facility) pair, compute the **cheapest insertion cost** --- the minimum additional distance to insert that demand into the facility's tour.
4.  Convert insertion distances to fuel cost: $\text{insertion\_fuel} = \text{insertion\_km} / \text{kml} \times \text{fuel\_price}$.
5.  Update: $\text{cost\_new} = \text{cost\_original} + \text{peso\_tsp} \times \text{insertion\_fuel}$.
6.  Repeat until assignments stabilize or `max_iter_tsp` is reached.

Because `insertion_fuel` is in the same monetary units as `cost_matrix`, `weight_tsp` is dimensionless: a value of 1 means the routing penalty is weighted equally with the original transport cost.

### Setup

The TSP loop needs a full distance matrix covering all demand points and facilities. We also pass `kml` and `fuel_price` so insertion distances (km) become comparable to the monetary cost matrix.

```{r tsp-setup}
# Full distance matrix: rows/cols 1:n_demand are demands, (n_demand+1):(n_demand+n_fac) are facilities
all_coords <- as.data.frame(rbind(
  st_coordinates(demand), st_coordinates(facilities)
))
all_points <- st_as_sf(all_coords, coords = c("X", "Y"))
dm_full <- distance_matrix(all_points)
```

### Solving with routing penalty

```{r tsp-solve}
tsp_args <- list(
  demand, facilities,
  weight_col = "workload",
  cost_matrix = cost_mat,
  facility_cost_col = "fixed_cost",
  worker_cost = worker_cost,
  worker_capacity = worker_capacity,
  max_workers_col = "max_workers",
  weight_tsp = 1,
  distance_matrix_full = dm_full,
  kml = kml,
  fuel_price = fuel_cost
)

# weight_tsp = 1: warm and cold
orce_tsp_result <- do.call(spopt::orce, tsp_args)
orce_tsp_meta <- attr(orce_tsp_result, "spopt")

orce_tsp_cold_result <- do.call(spopt::orce, c(tsp_args, warm_start = FALSE))
orce_tsp_cold_meta <- attr(orce_tsp_cold_result, "spopt")

# weight_tsp = 10: warm and cold
tsp_args10 <- modifyList(tsp_args, list(weight_tsp = 10))

orce_tsp10_result <- do.call(spopt::orce, tsp_args10)
orce_tsp10_meta <- attr(orce_tsp10_result, "spopt")

orce_tsp10_cold_result <- do.call(spopt::orce, c(tsp_args10, warm_start = FALSE))
orce_tsp10_cold_meta <- attr(orce_tsp10_cold_result, "spopt")

# weight_tsp = 100: warm and cold
tsp_args100 <- modifyList(tsp_args, list(weight_tsp = 100))

orce_tsp100_result <- do.call(spopt::orce, tsp_args100)
orce_tsp100_meta <- attr(orce_tsp100_result, "spopt")

orce_tsp100_cold_result <- do.call(spopt::orce, c(tsp_args100, warm_start = FALSE))
orce_tsp100_cold_meta <- attr(orce_tsp100_cold_result, "spopt")
```

### Comparing with and without TSP

The `Tour cost` column shows the total fuel cost of the TSP tours (all facilities combined). This is the actual routing cost when workers follow optimized tours instead of making independent round trips. The `Total + Tour` column adds this to the base objective.

```{r tsp-comparison}
# Convert tour distances to fuel cost
tour_fuel <- function(meta) {
  if (is.null(meta$total_tour_distance)) return(NA_real_)
  meta$total_tour_distance / kml * fuel_cost
}

tsp_comparison <- data.frame(
  Model = c("orce()", "orce(weight_tsp=1)", "orce(weight_tsp=1, cold)",
            "orce(weight_tsp=10)", "orce(weight_tsp=10, cold)",
            "orce(weight_tsp=100)", "orce(weight_tsp=100, cold)"),
  Facilities = c(orce_meta$n_selected, orce_tsp_meta$n_selected,
                 orce_tsp_cold_meta$n_selected,
                 orce_tsp10_meta$n_selected, orce_tsp10_cold_meta$n_selected,
                 orce_tsp100_meta$n_selected, orce_tsp100_cold_meta$n_selected),
  Transport = round(c(orce_meta$transport_cost, orce_tsp_meta$transport_cost,
                       orce_tsp_cold_meta$transport_cost,
                       orce_tsp10_meta$transport_cost, orce_tsp10_cold_meta$transport_cost,
                       orce_tsp100_meta$transport_cost, orce_tsp100_cold_meta$transport_cost), 1),
  Fixed = round(c(orce_meta$facility_cost, orce_tsp_meta$facility_cost,
                   orce_tsp_cold_meta$facility_cost,
                   orce_tsp10_meta$facility_cost, orce_tsp10_cold_meta$facility_cost,
                   orce_tsp100_meta$facility_cost, orce_tsp100_cold_meta$facility_cost), 1),
  Workers = round(c(orce_meta$worker_cost_total, orce_tsp_meta$worker_cost_total,
                     orce_tsp_cold_meta$worker_cost_total,
                     orce_tsp10_meta$worker_cost_total, orce_tsp10_cold_meta$worker_cost_total,
                     orce_tsp100_meta$worker_cost_total, orce_tsp100_cold_meta$worker_cost_total), 1),
  `Tour cost` = round(c(NA, tour_fuel(orce_tsp_meta), tour_fuel(orce_tsp_cold_meta),
                         tour_fuel(orce_tsp10_meta), tour_fuel(orce_tsp10_cold_meta),
                         tour_fuel(orce_tsp100_meta), tour_fuel(orce_tsp100_cold_meta)), 1),
  Iterations = c(orce_meta$tsp_iterations, orce_tsp_meta$tsp_iterations,
                  orce_tsp_cold_meta$tsp_iterations,
                  orce_tsp10_meta$tsp_iterations, orce_tsp10_cold_meta$tsp_iterations,
                  orce_tsp100_meta$tsp_iterations, orce_tsp100_cold_meta$tsp_iterations),
  Converged = c(NA, orce_tsp_meta$tsp_converged, orce_tsp_cold_meta$tsp_converged,
                orce_tsp10_meta$tsp_converged, orce_tsp10_cold_meta$tsp_converged,
                orce_tsp100_meta$tsp_converged, orce_tsp100_cold_meta$tsp_converged),
  `Time (s)` = round(c(orce_meta$solve_time, orce_tsp_meta$solve_time,
                        orce_tsp_cold_meta$solve_time,
                        orce_tsp10_meta$solve_time, orce_tsp10_cold_meta$solve_time,
                        orce_tsp100_meta$solve_time, orce_tsp100_cold_meta$solve_time), 3),
  check.names = FALSE
)
tsp_comparison$Total <- tsp_comparison$Transport + tsp_comparison$Fixed + tsp_comparison$Workers
tsp_comparison |> tt()
```

### What changed

The plot below highlights only the demand points whose assignment changed between `orce()` and `orce(weight_tsp = 1)`. For each reassigned point, a red segment shows the old assignment and a blue segment shows the new one. Grey dots are demands that stayed the same; triangles mark open facilities.

```{r tsp-plot, fig.width = 7, fig.height = 7}
demand_coords <- st_coordinates(demand)
fac_coords <- st_coordinates(facilities)

assign_base <- orce_result$demand$.facility
assign_tsp <- orce_tsp_result$demand$.facility
changed <- which(assign_base != assign_tsp)

n_changed <- length(changed)

plot(demand_coords, pch = 16, cex = 0.5, col = "grey70",
     xlab = "x (km)", ylab = "y (km)",
     main = sprintf("Reassigned demands: %d / %d", n_changed, n_demand))

# Segments for changed demands: red = old, blue = new
for (i in changed) {
  segments(demand_coords[i, 1], demand_coords[i, 2],
           fac_coords[assign_base[i], 1], fac_coords[assign_base[i], 2],
           col = adjustcolor("firebrick", alpha.f = 0.4), lwd = 1.2)
  segments(demand_coords[i, 1], demand_coords[i, 2],
           fac_coords[assign_tsp[i], 1], fac_coords[assign_tsp[i], 2],
           col = adjustcolor("steelblue", alpha.f = 0.5), lwd = 1.2)
}

# Highlight changed demands
points(demand_coords[changed, , drop = FALSE], pch = 16, cex = 0.9, col = "black")

# Open facilities (union of both solutions)
open_all <- sort(union(
  which(orce_result$facilities$.selected),
  which(orce_tsp_result$facilities$.selected)
))
points(fac_coords[open_all, , drop = FALSE], pch = 17, cex = 1.5, col = "grey30")

legend("topright", legend = c("old (weight_tsp=0)", "new (weight_tsp=1)", "unchanged"),
       col = c("firebrick", "steelblue", "grey70"),
       lwd = c(1.5, 1.5, NA), pch = c(NA, NA, 16), pt.cex = c(NA, NA, 0.8),
       bty = "n", cex = 0.85)
```

## When to use which

```{r when-table, echo = FALSE}
data.frame(
  Function = c("`cflp()`", "`orce()`", "`orce(weight_tsp > 0)`", "`orce::orce()`"),
  `Use when` = c(
    "Worker costs are negligible or already folded into fixed costs",
    "Worker allocation matters --- salary is a significant cost component",
    "Routing quality matters --- nearby demands should stay together",
    "You need the full IBGE model (multi-period, daily allowances)"
  ),
  check.names = FALSE
) |> tt()
```