---
title: "Warehouse location with orce"
format: html
editor: visual
---

```{r setup, include = FALSE}
library(spopt)
library(tinytable)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

The `orce()` function solves the warehouse location problem: given demand points and candidate facilities, decide which facilities to open, how to assign demand, and how many workers to place at each facility --- all while minimizing total cost.

This vignette compares three approaches to the same problem:

1.  **`cflp()`** --- Capacitated Facility Location: minimizes transport + fixed costs, treating capacity as a hard constraint
2.  **`orce()`** --- Warehouse Location: minimizes transport + fixed + worker costs, with worker count as a decision variable
3.  **`orce::orce()`** --- The reference implementation from the [orce](https://github.com/orce-ibge/orce) package

## A synthetic problem

We create 200 demand points (collection units) and 30 candidate facilities (agencies). Each demand point has a workload (collection days), and each facility has a fixed opening cost and a maximum number of workers.

```{r problem-setup}
library(sf)

set.seed(2024)

n_demand <- 200
n_fac <- 30

# Demand points: scattered across a 500x500 km area
demand <- st_as_sf(data.frame(
  id = seq_len(n_demand),
  x = runif(n_demand, 0, 500),
  y = runif(n_demand, 0, 500),
  workload = as.integer(rpois(n_demand, 15) + 5)
), coords = c("x", "y"))

# Facilities: 30 candidates at random locations
fac_x <- runif(n_fac, 10, 490)
fac_y <- runif(n_fac, 10, 490)
facilities <- st_as_sf(data.frame(
  id = seq_len(n_fac),
  x = fac_x,
  y = fac_y,
  fixed_cost = as.integer(runif(n_fac, 300, 800)),
  max_workers = as.integer(sample(3:8, n_fac, replace = TRUE))
), coords = c("x", "y"))
```

### Cost parameters

We model transport cost as fuel cost for a round trip:

$$\text{transport\_cost}_{ij} = \text{workload}_i \times \frac{2 \times \text{distance}_{ij}}{\text{kml}} \times \text{fuel\_cost}$$

Each worker can handle up to 60 collection days. Worker salary is 1500 per worker.

```{r cost-params}
kml <- 10              # km per liter
fuel_cost <- 6         # cost per liter
worker_capacity <- 60  # max workload per worker
worker_cost <- 1500    # salary per worker

# Distance matrix (Euclidean, treating coordinates as km)
dist_mat <- distance_matrix(demand, facilities)

# Per-unit transport cost: cost of one round trip to facility j from demand i
# (Used by cflp, which multiplies by weight internally)
unit_cost_mat <- 2 * dist_mat / kml * fuel_cost

# Total transport cost per demand-facility pair (includes workload weighting)
# (Used by orce, which does NOT multiply by weight)
cost_mat <- outer(demand$workload, rep(1, n_fac)) * unit_cost_mat
```

## Approach 1: CFLP

`cflp()` minimizes transport + fixed facility costs. It doesn't know about workers --- capacity is simply `max_workers * worker_capacity`. Because worker costs are not in the objective, it may open facilities that are cheap to open but expensive to staff.

```{r cflp-solve}
# CFLP needs a single capacity column
facilities$capacity <- facilities$max_workers * worker_capacity

# cflp() multiplies weight * cost internally, so we pass the per-unit cost
cflp_result <- cflp(
  demand, facilities,
  n_facilities = 0,
  weight_col = "workload",
  capacity_col = "capacity",
  facility_cost_col = "fixed_cost",
  cost_matrix = unit_cost_mat
)

cflp_meta <- attr(cflp_result, "spopt")
cflp_selected <- which(cflp_result$facilities$.selected)
```

## Approach 2: spopt::orce()

`orce()` adds worker costs to the objective. The solver decides how many workers each open facility needs, trading off transport savings against workforce costs.

```{r orce-solve}
orce_result <- orce(
  demand, facilities,
  weight_col = "workload",
  cost_matrix = cost_mat,
  facility_cost_col = "fixed_cost",
  worker_cost = worker_cost,
  worker_capacity = worker_capacity,
  max_workers_col = "max_workers"
)

orce_meta <- attr(orce_result, "spopt")
orce_selected <- which(orce_result$facilities$.selected)
```

Workers per open facility:

```{r orce-workers}
open <- orce_result$facilities[orce_result$facilities$.selected, ]
data.frame(
  facility = open$id,
  workers = open$.workers,
  assigned = open$.n_assigned,
  utilization = round(open$.utilization, 2)
) |> tt()
```

## Approach 3: orce::orce() (reference)

The original orce package solves the same MIP using R's ROI infrastructure. We set up the identical problem to verify our Rust implementation matches.

```{r orce-pkg, eval = requireNamespace("orce", quietly = TRUE)}
library(orce)

uc_ids <- paste0("uc_", seq_len(n_demand))
ag_ids <- paste0("ag_", seq_len(n_fac))

ucs_df <- data.frame(
  uc = uc_ids,
  agencia_codigo = ag_ids[1],
  dias_coleta = demand$workload,
  viagens = 1L,
  data = "2024-01",
  diaria_valor = 0,
  stringsAsFactors = FALSE
)

agencias_df <- data.frame(
  agencia_codigo = ag_ids,
  n_entrevistadores_agencia_max = facilities$max_workers,
  custo_fixo = facilities$fixed_cost,
  diaria_valor = 0,
  stringsAsFactors = FALSE
)

distancias_ucs_df <- expand.grid(
  uc = uc_ids,
  agencia_codigo = ag_ids,
  stringsAsFactors = FALSE
)
distancias_ucs_df$distancia_km <- as.vector(dist_mat)
distancias_ucs_df$duracao_horas <- 0
distancias_ucs_df$diaria_municipio <- FALSE
distancias_ucs_df$diaria_pernoite <- FALSE

orce_pkg_args <- list(
  ucs = ucs_df,
  agencias = agencias_df,
  distancias_ucs = distancias_ucs_df,
  dias_coleta_entrevistador_max = worker_capacity,
  remuneracao_entrevistador = worker_cost,
  n_entrevistadores_min = 1L,
  custo_litro_combustivel = fuel_cost,
  custo_hora_viagem = 0,
  kml = kml,
  dias_treinamento = 0,
  peso_tsp = 0,
  adicional_troca_jurisdicao = 0,
  rel_tol = 0,
  use_cache = FALSE
)

# Integer workers (default)
orce_pkg_time <- system.time({
  orce_pkg_result <- do.call(orce::orce, orce_pkg_args)
})

# Continuous workers (LP relaxation of w[j])
orce_pkg_cont_time <- system.time({
  orce_pkg_cont_result <- do.call(orce::orce,
    c(orce_pkg_args, n_entrevistadores_tipo = "continuous"))
})

orce_pkg_obj <- attr(orce_pkg_result, "valor")
orce_pkg_cont_obj <- attr(orce_pkg_cont_result, "valor")
```

## Comparison

```{r comparison-table}
# CFLP transport cost = objective - fixed costs
cflp_transport <- cflp_meta$objective - sum(facilities$fixed_cost[cflp_selected])
cflp_fixed <- sum(facilities$fixed_cost[cflp_selected])

# Compute CFLP's worker cost post-hoc (not part of its optimization)
cflp_workers_per_fac <- ceiling(
  tapply(demand$workload[cflp_result$demand$.facility],
         cflp_result$demand$.facility,
         sum) / worker_capacity
)
cflp_worker_total <- sum(cflp_workers_per_fac) * worker_cost

has_orce_pkg <- exists("orce_pkg_result")

comparison <- data.frame(
  Model = c("cflp()", "orce()",
            if (has_orce_pkg) c("orce::orce(integer)", "orce::orce(continuous)")),
  Facilities = c(
    sum(cflp_result$facilities$.selected),
    orce_meta$n_selected,
    if (has_orce_pkg) c(
      nrow(orce_pkg_result$resultado_agencias_otimo),
      nrow(orce_pkg_cont_result$resultado_agencias_otimo)
    )
  ),
  Transport = round(c(
    cflp_transport,
    orce_meta$transport_cost,
    if (has_orce_pkg) c(
      sum(orce_pkg_result$resultado_ucs_otimo$custo_deslocamento),
      sum(orce_pkg_cont_result$resultado_ucs_otimo$custo_deslocamento)
    )
  ), 1),
  Fixed = round(c(
    cflp_fixed,
    orce_meta$facility_cost,
    if (has_orce_pkg) c(
      sum(orce_pkg_result$resultado_agencias_otimo$custo_fixo),
      sum(orce_pkg_cont_result$resultado_agencias_otimo$custo_fixo)
    )
  ), 1),
  Workers = round(c(
    cflp_worker_total,
    orce_meta$worker_cost_total,
    if (has_orce_pkg) c(
      sum(orce_pkg_result$resultado_agencias_otimo$custo_total_entrevistadores),
      sum(orce_pkg_cont_result$resultado_agencias_otimo$custo_total_entrevistadores)
    )
  ), 1),
  `Time (s)` = round(c(
    cflp_meta$solve_time,
    orce_meta$solve_time,
    if (has_orce_pkg) c(
      orce_pkg_time[["elapsed"]],
      orce_pkg_cont_time[["elapsed"]]
    )
  ), 3),
  check.names = FALSE
)
comparison$Total <- comparison$Transport + comparison$Fixed + comparison$Workers
comparison |> tt()
```

The key insight: `cflp()` and `orce()` may select different facilities because `orce()` accounts for workforce costs during optimization, while `cflp()` only considers transport and fixed costs. When worker costs are significant (as they often are), the `orce()` solution is more cost-effective overall.

## When to use which

```{r when-table, echo = FALSE}
data.frame(
  Function = c("`cflp()`", "`orce()`", "`orce::orce()`"),
  `Use when` = c(
    "Worker costs are negligible or already folded into fixed costs",
    "Worker allocation matters --- salary is a significant cost component",
    "You need the full IBGE model (multi-period, daily allowances, TSP)"
  ),
  check.names = FALSE
) |> tt()
```
