---
title: "Route optimization with spopt"
format: html
editor: visual
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

*Route optimization* asks: given a set of stops that need to be visited, what's the best sequence? This is the Traveling Salesman Problem (TSP) when you have one vehicle, and the Vehicle Routing Problem (VRP) when you have a fleet. These problems show up everywhere - delivery routing, field service scheduling, sales territory coverage, equipment inspection rounds.

spopt's routing functions take a travel-time matrix (from r5r, OSRM, mapboxapi, or any routing engine) and find the optimal stop sequence. The solvers handle real-world constraints: open-ended routes, fixed start/end points, vehicle capacities, and time windows.

## Setting up the problem

Route optimization has three components:

1.  **Stops**: Locations that need to be visited
2.  **Depot**: Where vehicles start (and possibly return)
3.  **Cost matrix**: Travel time or distance between every pair of stops

The cost matrix is the key input. Euclidean distance works for quick prototyping, but real routing needs real travel times - a stop across a river might be 2 miles as the crow flies but 15 minutes by car. That's where an external routing engine comes in.

Let's work through a realistic scenario: a delivery station in west Fort Worth dispatching drivers to 25 residential addresses across Fort Worth, Benbrook, and Aledo.

```{r}
library(spopt)
library(tidyverse)
library(sf)
library(mapgl)

# Load the bundled delivery data
data(delivery_data)

stops <- delivery_data$stops
ttm <- delivery_data$matrix

# Take a look at what we have
stops |>
  st_drop_geometry() |>
  select(id, address, packages)
```

The `delivery_data` dataset includes 26 locations: an Amazon delivery station (the depot) and 25 residential stops across western Tarrant and Parker counties. Each stop has a package count, and `ttm` is a 26x26 travel-time matrix in minutes computed with r5r from an OpenStreetMap road network.

```{r}
# Depot is the first row
depot <- stops |> filter(id == "depot")
deliveries <- stops |> filter(id != "depot")

maplibre(style = openfreemap_style("bright"), bounds = stops) |>
  add_circle_layer(
    id = "deliveries",
    source = deliveries,
    circle_color = "steelblue",
    circle_radius = 6,
    circle_stroke_color = "white",
    circle_stroke_width = 2,
    tooltip = "address"
  ) |>
  add_circle_layer(
    id = "depot",
    source = depot,
    circle_color = "#e41a1c",
    circle_radius = 10,
    circle_stroke_color = "white",
    circle_stroke_width = 3,
    tooltip = "address"
  )
```

The red marker is the depot (Amazon DDA9 on San Jacinto Dr). Blue markers are delivery stops spread from central Fort Worth west into Aledo and Willow Park. Some stops cluster tightly in Ridglea and Benbrook; others are 20+ minutes out in Parker County. That geographic spread is what makes the stop sequence matter.

## Generating the travel-time matrix

The bundled matrix was built with r5r using the same workflow described in the [travel-time matrices vignette](travel-time-matrices.html). The key difference for routing is that you need a **square** matrix - every stop to every other stop - rather than the rectangular demand-to-facility matrix used in facility location.

```{r}
#| eval: false
# How the matrix was generated (requires Java 21 + OSM data)
library(r5r)
options(java.parameters = "-Xmx4G")
rJavaEnv::java_quick_install(version = 21)

r5r_core <- build_network(data_path = "path/to/osm/directory")

# Prepare points for r5r
r5r_pts <- stops |>
  st_coordinates() |>
  as_tibble() |>
  rename(lon = X, lat = Y) |>
  mutate(id = stops$id)

# Square matrix: all stops to all stops
ttm_long <- travel_time_matrix(
  r5r_core,
  origins = r5r_pts,
  destinations = r5r_pts,
  mode = "CAR",
  departure_datetime = as.POSIXct("2025-03-15 08:00:00"),
  max_trip_duration = 120
)

# Reshape to matrix
ttm <- ttm_long |>
  select(from_id, to_id, travel_time_p50) |>
  pivot_wider(names_from = to_id, values_from = travel_time_p50) |>
  arrange(match(from_id, stops$id)) |>
  select(-from_id) |>
  as.matrix()
```

Any routing engine that produces pairwise travel times works here - OSRM, Valhalla, mapboxapi, Google Distance Matrix. The optimizer doesn't care how the matrix was computed; it just needs the numbers.

## Single-driver route: TSP

The simplest case: one driver, all 25 stops, return to the depot. The `route_tsp()` function finds the shortest sequence.

```{r}
result <- route_tsp(stops, start = 1, cost_matrix = ttm)

meta <- attr(result, "spopt")
cat(sprintf("Optimized route: %.0f minutes\n", meta$total_cost))
cat(sprintf("Nearest-neighbor baseline: %.0f minutes\n", meta$nn_cost))
cat(sprintf("Improvement: %.1f%%\n", meta$improvement_pct))
```

The solver starts with a nearest-neighbor heuristic (always go to the closest unvisited stop), then improves it with 2-opt and or-opt local search. The improvement over nearest-neighbor is typical - greedy heuristics get you close, but the local search squeezes out the remaining inefficiency.

Let's look at the visit order:

```{r}
result |>
  st_drop_geometry() |>
  filter(.visit_order <= 10) |>
  arrange(.visit_order) |>
  select(id, address, .visit_order)
```

The optimizer clusters geographically - it handles the Aledo stops together rather than bouncing back and forth. That's the whole point. Let's see it on a map.

The bundled `delivery_data` includes pre-computed route geometries from r5r, so we can draw the actual driving paths:

```{r}
tsp_route <- delivery_data$tsp_route

# Number stops by visit order
result_ordered <- result |>
  filter(!is.na(.visit_order)) |>
  mutate(label = as.character(.visit_order))

maplibre(style = openfreemap_style("bright"), bounds = stops) |>
  add_line_layer(
    id = "route",
    source = tsp_route,
    line_color = "#2563eb",
    line_width = 3,
    line_opacity = 0.8
  ) |>
  add_circle_layer(
    id = "stops",
    source = result_ordered |> filter(id != "depot"),
    circle_color = "#2563eb",
    circle_radius = 7,
    circle_stroke_color = "white",
    circle_stroke_width = 2,
    tooltip = "address"
  ) |>
  add_symbol_layer(
    id = "labels",
    source = result_ordered |> filter(id != "depot"),
    text_field = get_column("label"),
    text_size = 11,
    text_color = "white",
    text_halo_color = "#1e40af",
    text_halo_width = 1.5
  ) |>
  add_circle_layer(
    id = "depot",
    source = depot,
    circle_color = "#dc2626",
    circle_radius = 10,
    circle_stroke_color = "white",
    circle_stroke_width = 3,
    tooltip = "address"
  )
```

The route follows actual roads between stops. Notice how it sweeps through the western stops (Aledo, Willow Park) in one pass rather than zigzagging back to central Fort Worth.

### Open routes

Drivers don't always return to the depot. A field technician might start at the office and end at their last appointment. Set `end = NULL` for an open route:

```{r}
result_open <- route_tsp(stops, start = 1, end = NULL, cost_matrix = ttm)
meta_open <- attr(result_open, "spopt")

cat(sprintf("Closed route: %.0f minutes\n", meta$total_cost))
cat(sprintf("Open route:   %.0f minutes\n", meta_open$total_cost))
cat(sprintf("Saved by not returning: %.0f minutes\n",
            meta$total_cost - meta_open$total_cost))
```

Dropping the return leg saves real time, especially when the last stop is far from the depot.

### Fixed start and end

Sometimes the start and end points are different - a courier picks up from a warehouse and must end at a specific drop-off location. Use `start` and `end` to fix both endpoints:

```{r}
# Start at depot (1), end at the farthest delivery (index of a Parker County stop)
parker_stop <- which(grepl("Willow Park|Hudson Oaks|Aledo", stops$address))[1]

result_path <- route_tsp(stops, start = 1, end = parker_stop, cost_matrix = ttm)
meta_path <- attr(result_path, "spopt")

cat(sprintf("Route type: %s\n", meta_path$route_type))
cat(sprintf("Total time: %.0f minutes\n", meta_path$total_cost))
```

## Fleet routing: VRP

Real delivery operations have multiple drivers, and each van has a capacity limit. The `route_vrp()` function solves the capacitated vehicle routing problem: assign stops to vehicles and sequence each vehicle's route, all while respecting capacity constraints.

Our stops have package counts ranging from 3 to 5. With a van capacity of 35 packages:

```{r}
cat(sprintf("Total packages: %d\n", sum(stops$packages)))
cat(sprintf("Van capacity: 35\n"))
cat(sprintf("Minimum vans needed: %d\n", ceiling(sum(stops$packages) / 35)))
```

```{r}
result_vrp <- route_vrp(
  stops,
  depot = 1,
  demand_col = "packages",
  vehicle_capacity = 35,
  cost_matrix = ttm
)

meta_vrp <- attr(result_vrp, "spopt")
cat(sprintf("Vehicles used: %d\n", meta_vrp$n_vehicles))
cat(sprintf("Total drive time: %.0f minutes\n", meta_vrp$total_cost))
```

The solver uses a Clarke-Wright savings heuristic for initial route construction, then improves with intra-route 2-opt (resequencing stops within each route) and inter-route moves (swapping stops between vehicles).

Let's see how the stops got assigned:

```{r}
for (v in seq_len(meta_vrp$n_vehicles)) {
  vehicle_stops <- result_vrp |>
    filter(.vehicle == v) |>
    arrange(.visit_order)
  load <- meta_vrp$vehicle_loads[v]
  time <- meta_vrp$vehicle_costs[v]
  cat(sprintf("\nVan %d: %d stops, %d packages, %.0f min drive time\n",
              v, nrow(vehicle_stops), load, time))
}
```

### Visualizing vehicle routes

The bundled data includes road geometries for each vehicle's route. Let's see how the fleet covers the delivery area:

```{r}
vrp_route <- delivery_data$vrp_route |>
  mutate(vehicle = as.character(vehicle))

vrp_stops <- result_vrp |>
  filter(.vehicle > 0) |>
  mutate(.vehicle = as.character(.vehicle))

vehicle_colors <- c("#e41a1c", "#377eb8", "#4daf4a")
vehicle_ids <- as.character(seq_len(meta_vrp$n_vehicles))

maplibre(style = openfreemap_style("bright"), bounds = stops) |>
  add_line_layer(
    id = "routes",
    source = vrp_route,
    line_color = match_expr(
      column = "vehicle",
      values = vehicle_ids,
      stops = vehicle_colors[seq_along(vehicle_ids)]
    ),
    line_width = 3,
    line_opacity = 0.8
  ) |>
  add_circle_layer(
    id = "stops",
    source = vrp_stops,
    circle_color = match_expr(
      column = ".vehicle",
      values = vehicle_ids,
      stops = vehicle_colors[seq_along(vehicle_ids)]
    ),
    circle_radius = 7,
    circle_stroke_color = "white",
    circle_stroke_width = 2,
    tooltip = "address"
  ) |>
  add_circle_layer(
    id = "depot",
    source = depot,
    circle_color = "black",
    circle_radius = 10,
    circle_stroke_color = "white",
    circle_stroke_width = 3,
    tooltip = "address"
  )
```

Each color is a different van's route. The solver clusters geographically - nearby stops go on the same truck - while respecting the 35-package capacity. All three routes start and end at the depot (black marker).

### Constraining the fleet size

If you have exactly 2 vans available, set `n_vehicles`:

```{r}
result_2vans <- route_vrp(
  stops,
  depot = 1,
  demand_col = "packages",
  vehicle_capacity = 50,
  n_vehicles = 2,
  cost_matrix = ttm
)

meta_2 <- attr(result_2vans, "spopt")
cat(sprintf("Vehicles: %d\n", meta_2$n_vehicles))
cat(sprintf("Total time: %.0f min (vs %.0f min with %d vans)\n",
            meta_2$total_cost, meta_vrp$total_cost, meta_vrp$n_vehicles))
```

Fewer vehicles means longer routes per driver. The solver finds the best assignment given the constraint, but you'll want to check that the resulting route times are realistic for a single shift.

### Shift limits and service time

If drivers have a 90-minute shift limit, set `max_route_time`. The solver will split routes so that no vehicle exceeds the time budget. You can also account for time spent at each stop with `service_time` -- loading/unloading, signatures, etc.

```{r}
result_shift <- route_vrp(
  stops,
  depot = 1,
  demand_col = "packages",
  vehicle_capacity = 50,
  cost_matrix = ttm,
  service_time = 3,       # 3 minutes at each stop
  max_route_time = 90     # 90-minute shift limit
)

meta_shift <- attr(result_shift, "spopt")
cat(sprintf("Vehicles needed: %d\n", meta_shift$n_vehicles))
for (v in seq_len(meta_shift$n_vehicles)) {
  cat(sprintf("  Van %d: %.0f min drive + %.0f min service = %.0f min total\n",
              v, meta_shift$vehicle_costs[v],
              meta_shift$vehicle_times[v] - meta_shift$vehicle_costs[v],
              meta_shift$vehicle_times[v]))
}
```

`max_route_time` is a hard constraint: total time (travel + service) on every route stays within the limit. The objective is still minimizing total travel time, so the solver won't use extra vehicles unless the time constraint forces it.

### Balancing route times

The cost optimizer finds the cheapest set of routes, but that can produce uneven workloads -- one driver finishes in 30 minutes while another is out for 80. Setting `balance = "time"` runs a post-optimization phase that redistributes stops to reduce the longest route time, at the cost of a small bounded cost increase.

```{r}
result_balanced <- route_vrp(
  stops,
  depot = 1,
  demand_col = "packages",
  vehicle_capacity = 50,
  cost_matrix = ttm,
  service_time = 3,
  balance = "time"
)

meta_bal <- attr(result_balanced, "spopt")
for (v in seq_len(meta_bal$n_vehicles)) {
  cat(sprintf("  Van %d: %.0f min total\n", v, meta_bal$vehicle_times[v]))
}
cat(sprintf("Cost without balancing: %.0f min\n", meta_vrp$total_cost))
cat(sprintf("Cost with balancing:    %.0f min\n", meta_bal$total_cost))
```

The balancing phase only runs when `method = "2-opt"` (the default). It targets the longest route and tries moving or swapping stops to shorten it. Capacity and time constraints are still respected.

## TSP vs VRP: when to use which

| Scenario | Function | Key parameters |
|------------------------|------------------------|------------------------|
| One driver, return to base | `route_tsp(start = 1)` | Default closed tour |
| One driver, end anywhere | `route_tsp(start = 1, end = NULL)` | Open route |
| One driver, fixed endpoint | `route_tsp(start = 1, end = 5)` | Path between two points |
| Multiple drivers, capacity limits | `route_vrp(demand_col, vehicle_capacity)` | Fleet optimization |
| Fixed fleet size | `route_vrp(n_vehicles = 3)` | Constrained fleet |
| Shift time limits | `route_vrp(max_route_time = 90)` | Per-route duration cap |

For most delivery and field service problems, start with `route_tsp()` to understand the single-vehicle baseline, then move to `route_vrp()` when you need to split work across a fleet.

## Next steps

-   [Travel-Time Cost Matrices](travel-time-matrices.html) - Generate travel-time matrices with r5r
-   [Facility Location](facility-location.html) - Optimize depot and warehouse placement
-   [Getting Started](getting-started.html) - Overview of all spopt capabilities