---
title: "Zone allocation with allocate_zones"
format: html
editor: visual
bibliography: references.bib
params:
  n_sample: 50
---

```{r setup, include = FALSE}
library(spopt)
library(tinytable)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
options(tigris_use_cache = TRUE)
n_sample <- params$n_sample
spopt_enable_cache(persistent = TRUE)
```

`allocate_zones()` partitions geographic units into zones, each with a center, such that no unit exceeds `max_distance` from its center. It automatically discovers the minimum number of centers via LSCP, then assigns units using one of three methods:

-   **`"nearest"`**: Greedy nearest-center assignment (fastest)
-   **`"p_median"`**: Minimizes total weighted distance (optimal)
-   **`"cflp"`**: Adds capacity constraints (capped workloads)

This vignette demonstrates all three methods with TSP-based zone sequencing using Census tracts from Tarrant County, Texas.

## Setup

```{r data}
library(spopt)
library(tidycensus)
library(tidyverse)
library(sf)

tarrant <- get_acs(
  geography = "tract",
  variables = "B01003_001",
  state = "TX",
  county = "Tarrant",
  geometry = TRUE,
  year = 2023
) |>
  filter(estimate > 0) |>
  rename(population = estimate)

zones_sf <- tarrant |>
  st_centroid() |>
  select(GEOID, population)

# Sample for faster rendering
set.seed(42)
zones_sf <- zones_sf |> slice_sample(n = min(n_sample, nrow(zones_sf)))
tarrant <- tarrant |> filter(GEOID %in% zones_sf$GEOID)

dm <- units::drop_units(st_distance(zones_sf)) / 1000
nrow(zones_sf)
```

## Three methods compared

```{r solve}
result_near <- allocate_zones(zones_sf, max_distance = 5,
                              cost_matrix = dm, verbose = TRUE)

result_opt <- allocate_zones(zones_sf, max_distance = 5, method = "p_median",
                             weight_col = "population",
                             cost_matrix = dm, verbose = TRUE)

result_cap <- allocate_zones(zones_sf, max_distance = 5, method = "cflp",
                             capacity = 3,
                             cost_matrix = dm, verbose = TRUE)
```

```{r compare}
m1 <- attr(result_near, "spopt")
m2 <- attr(result_opt, "spopt")
m3 <- attr(result_cap, "spopt")

data.frame(
  Metric = c("Centers", "Total distance (km)", "Max tracts/center", "Solve time (s)"),
  nearest = c(
    m1$n_selected,
    round(sum(result_near$zones$.distance), 1),
    max(table(result_near$zones$.center)),
    round(m1$solve_time, 3)
  ),
  p_median = c(
    m2$n_selected,
    round(sum(result_opt$zones$.distance), 1),
    max(table(result_opt$zones$.center)),
    round(m2$solve_time, 3)
  ),
  `cflp(5)` = c(
    m3$n_selected,
    round(sum(result_cap$zones$.distance), 1),
    max(table(result_cap$zones$.center)),
    round(m3$solve_time, 3)
  ),
  check.names = FALSE
) |> tt()
```

## Mapping

```{r map, fig.width=14, fig.height=5, out.width="100%"}
map_near <- tarrant |>
  left_join(st_drop_geometry(result_near$zones) |>
              select(GEOID, zone = .center), by = "GEOID") |>
  mutate(method = "nearest")

map_opt <- tarrant |>
  left_join(st_drop_geometry(result_opt$zones) |>
              select(GEOID, zone = .center), by = "GEOID") |>
  mutate(method = "p_median")

map_cap <- tarrant |>
  left_join(st_drop_geometry(result_cap$zones) |>
              select(GEOID, zone = .center), by = "GEOID") |>
  mutate(method = "cflp (cap=5)")

bind_rows(map_near, map_opt, map_cap) |>
  mutate(method = factor(method, levels = c("nearest", "p_median", "cflp (cap=5)"))) |>
  ggplot() +
  geom_sf(aes(fill = factor(zone)), color = "white", linewidth = 0.1) +
  scale_fill_discrete(guide = "none") +
  facet_wrap(~method, ncol = 3) +
  theme_void()
```

## Zone sequencing

When `sequence = TRUE`, zones and tracts are ordered via TSP for geographic routing. Inter-zone distances use minimum pairwise distance between zone members (matching the surveyzones approach). Within each zone, tracts are oriented so the entry tract is closest to the previous zone's exit.

```{r sequence}
result_seq <- allocate_zones(zones_sf, max_distance = 5, method = "p_median",
                             weight_col = "population",
                             cost_matrix = dm, sequence = TRUE, verbose = TRUE)

# Show first 15 tracts in visit order
st_drop_geometry(result_seq$zones) |>
  select(GEOID, .center, .zone_order, .tract_order) |>
  arrange(.zone_order, .tract_order) |>
  head(15) |>
  tt()
```

```{r seqmap, fig.width=8, fig.height=6, out.width="100%"}
map_seq <- tarrant |>
  left_join(st_drop_geometry(result_seq$zones) |>
              select(GEOID, zone_order = .zone_order, tract_order = .tract_order,
                     zone_id = .zone_id),
            by = "GEOID")

# Build visit path as a line connecting centroids in sequence
centroids <- st_centroid(map_seq) |>
  arrange(zone_order, tract_order)
visit_line <- centroids |>
  summarise(geometry = st_cast(st_combine(geometry), "LINESTRING"))

ggplot() +
  geom_sf(data = map_seq, aes(fill = zone_order), color = "white", linewidth = 0.1) +
  geom_sf(data = visit_line, color = "black", linewidth = 0.3, alpha = 0.5) +
  geom_sf_text(data = map_seq, aes(label = zone_id), size = 1.8, color = "black") +
  scale_fill_viridis_c(option = "turbo", name = "Zone\norder") +
  labs(title = "Visit sequence (zone.tract labels, path line)") +
  theme_void()
```

Labels show `zone_order.tract_order` (e.g., `3.02` = zone 3, tract 2). The black line traces the full visit path across all zones in sequence.

## With partitions

```{r partition}
coords <- st_coordinates(zones_sf)
zones_sf$region <- ifelse(coords[, 1] < median(coords[, 1]), "west", "east")

result_part <- allocate_zones(zones_sf, max_distance = 5, method = "p_median",
                              partition_col = "region",
                              cost_matrix = dm, verbose = TRUE)

data.frame(
  Approach = c("Partitioned", "Unpartitioned"),
  Centers = c(attr(result_part, "spopt")$n_selected, m2$n_selected)
) |> tt()
```

## Method summary

```{r method-summary, echo = FALSE}
data.frame(
  ` ` = c("K discovery", "Assignment", "Weights", "Capacity", "Sequencing", "Best for"),
  nearest = c("LSCP", "Nearest (greedy)", "Not used", "No", "sequence = TRUE", "Speed"),
  p_median = c("LSCP", "P-median (optimal)", "Optional", "No", "sequence = TRUE", "Non-uniform weights"),
  cflp = c("LSCP + increment", "CFLP (optimal)", "Optional", "Yes (capacity)", "sequence = TRUE", "Capped workloads"),
  check.names = FALSE
) |> tt()
```